---
title: "Preparing i-ADHoRe input from GFF/GTF files"
author: "iadhoreR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)

# Work whether the package is installed or we are knitting from the source repo.
if (requireNamespace("iadhoreR", quietly = TRUE) &&
    nzchar(system.file("extdata", package = "iadhoreR"))) {
  library(iadhoreR)
} else {
  devtools::load_all(".")
}

# Locate a bundled example file (installed package, or inst/extdata in the repo).
ex <- function(f) {
  p <- system.file("extdata", f, package = "iadhoreR")
  if (nzchar(p)) p else file.path("inst", "extdata", f)
}
```

## What this shows

To run i-ADHoRe you need, per genome, one **gene list** file per chromosome:
a position-ordered list of gene identifiers, each tagged with its strand
(`AT1G01010+`). Those identifiers must match the IDs used in your BLAST/DIAMOND
homology table — which usually means matching the headers of the protein FASTA
the homology search was built from.

Real annotation files make this awkward: the gene ID you need can live under
any GFF attribute key (`ID`, `Name`, `protein_id`, `Source_ID`, …), feature
types like `CDS` repeat each gene over many rows, and some features carry no
strand. This document walks through the three helpers that handle all of that:

1. `inspect_gff()` — survey feature types and attribute keys.
2. `recommend_parse_gff()` — pick the right `feature_type` / `id_attribute` by
   matching against a proteome.
3. `parse_gff()` — write the per-chromosome gene lists.

The examples use the small genome subsets bundled with the package.

## 1. Inspecting a GFF

`inspect_gff()` reports which feature types exist and which attribute keys each
one carries, with example IDs — so you can see what you have before committing
to any parameters.

```{r inspect}
ath_gff <- ex("annotation.selected_transcript.all_features.ath.gff3")
info <- inspect_gff(ath_gff)
```

The `mRNA` rows carry an `ID` like `AT1G01010.1`, which is exactly how the
*Arabidopsis* proteome headers are written.

## 2. Matching a GFF to a proteome

You rarely have to guess the parameters. `recommend_parse_gff()` tries **every**
feature-type / attribute-key combination present in the file and reports the one
whose IDs best overlap your protein FASTA.

```{r recommend-ath}
ath_pep <- ex("proteome.selected_transcript.ath.fasta")
rec_ath <- recommend_parse_gff(ath_gff, ath_pep)
```

Here it lands on `feature_type = "mRNA"`, `id_attribute = "ID"` — the
transcript-level identifiers.

The right answer depends on how the proteome was keyed, not on a fixed
convention. The bundled *Aristolochia fimbriata* proteome uses **gene-level**
IDs (`>Af01G000001`), so the same function recommends a different feature type
for it:

```{r recommend-aristolochia}
afi_gff <- ex("Aristolochia_fimbriata.gff")
afi_pep <- ex("Aristolochia_fimbriata.fasta")
rec_afi <- recommend_parse_gff(afi_gff, afi_pep)
rec_afi[c("feature_type", "id_attribute", "n_fasta", "n_matched")]
```

Because the recommendation considers **all** attribute keys (not a hardcoded
short list), it also handles annotations that stash the matching ID in
non-standard places. The small synthetic GFF below keeps the FASTA IDs under a
`protein_id` key on `CDS` rows — a layout that defeats a fixed `ID`/`Name`
search but is found here automatically:

```{r recommend-nonstandard}
gff_lines <- c(
  "##gff-version 3",
  "chr1\tsrc\tgene\t100\t900\t.\t+\t.\tID=gene:G1",
  "chr1\tsrc\tCDS\t100\t300\t.\t+\t0\tID=c1;Parent=gene:G1;protein_id=PROT1",
  "chr1\tsrc\tCDS\t400\t900\t.\t+\t1\tID=c1;Parent=gene:G1;protein_id=PROT1",
  "chr1\tsrc\tgene\t2000\t2500\t.\t-\t.\tID=gene:G2",
  "chr1\tsrc\tCDS\t2000\t2500\t.\t-\t0\tID=c2;Parent=gene:G2;protein_id=PROT2"
)
tmp_gff <- tempfile(fileext = ".gff");  writeLines(gff_lines, tmp_gff)
tmp_pep <- tempfile(fileext = ".pep");  writeLines(c(">PROT1", "MAAA",
                                                     ">PROT2", "MBBB"), tmp_pep)

rec_ns <- recommend_parse_gff(tmp_gff, tmp_pep)
rec_ns[c("feature_type", "id_attribute", "n_matched")]
```

## 3. Writing the gene lists

Feed the recommended parameters into `parse_gff()`. It writes one `.lst` file
per chromosome and returns a named vector you can pass straight to
`write_iadhore_config()`.

```{r parse}
out_dir <- file.path(tempdir(), "ath_lists")
ath_lists <- parse_gff(
  ath_gff,
  output_dir   = out_dir,
  genome_name  = "ath",
  feature_type = rec_ath$feature_type,
  id_attribute = rec_ath$id_attribute,
  chromosomes  = rec_ath$chromosomes
)

length(ath_lists)            # one file per chromosome
head(ath_lists, 2)
# first few genes of the first chromosome, in position order with strand tags
head(readLines(ath_lists[[1]]), 5)
```

`parse_gff()` quietly does two things that matter on real data:

- **One entry per gene.** When the chosen feature type repeats a gene across
  rows (typical for `CDS` or `exon`), only the first occurrence — at the gene's
  lowest coordinate — is kept, so the list isn't padded with duplicates.
- **Strand-safe.** Features whose strand is `.` (unknown) can't be oriented in a
  gene list, so they are dropped with a warning rather than aborting the run.

Both are visible on the synthetic `CDS` example from above: two `CDS` rows for
`PROT1` collapse to a single ordered entry.

```{r parse-cds}
cds_dir <- file.path(tempdir(), "cds_lists")
cds_lists <- parse_gff(tmp_gff, output_dir = cds_dir, genome_name = "demo",
                       feature_type = "CDS", id_attribute = "protein_id")
readLines(cds_lists[["chr1"]])
```

## 4. When the IDs genuinely don't line up

Sometimes a proteome and an annotation simply use different identifier systems —
e.g. FASTA headers like `jgi|Chlin1|10000|g5205.t1` whose gene-model name never
appears in the GFF. Rather than silently matching a handful of genes,
`recommend_parse_gff()` stops and tells you *where* the IDs do correspond, so you
can reconcile them before going further.

```{r mismatch, error=TRUE}
# Same GFF as above, but a proteome whose primary IDs (SEQ_*) are not stored in
# the GFF; the real link is the second header field, which matches protein_id.
mismatch_pep <- tempfile(fileext = ".pep")
writeLines(c(">SEQ_0001 PROT1", "MAAA",
             ">SEQ_0002 PROT2", "MBBB"), mismatch_pep)

recommend_parse_gff(tmp_gff, mismatch_pep)
```

The message names the attribute (`protein_id`) that matches a FASTA header
sub-field, which tells you the fix: rename the FASTA headers, or rebuild the
homology table using the same IDs the GFF stores.

## Next step

Once every genome has its gene lists, assemble the run configuration:

```{r config, eval=FALSE}
write_iadhore_config(
  genomes     = list(ath = ath_lists, afi = afi_lists),
  blast_table = "blast_table.txt",
  output_path = "iadhore_out",
  file        = "iadhore.ini"
)
```

See the *Getting Started with iadhoreR* tutorial for the full collinearity
workflow from here.
```{r cleanup, include=FALSE}
unlink(c(out_dir, cds_dir), recursive = TRUE)
```
