---
title: "Getting Started with iadhoreR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with iadhoreR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**iadhoreR** is an R interface to
[i-ADHoRe 3.0](https://github.com/VIB-PSB/i-ADHoRe), a tool for detecting
collinear (syntenic) regions within and between genomes. It helps answer
questions like:

- Which chromosomal regions in species A are syntenic with regions in species B?
- Which genomic segments are remnants of ancient whole-genome duplications?

This tutorial walks through a complete analysis using two bundled example
species (Chr1 subsets included in the package; use your own full genomes for
real analyses):

- **ath** — *Arabidopsis thaliana* (Chr1: ~7,150 proteins)
- **vvi** — *Vitis vinifera* (chr1: ~1,400 proteins)

The typical workflow has five stages:

```
GFF + FASTA                    i-ADHoRe config
    │                               │
    ▼                               ▼
parse_gff()   run_diamond()    write_iadhore_config()
    │               │               │
    │               ▼               │
    │         blast_to_families()   │
    │               │               │
    └───────────────┴───────────────┘
                    │
                    ▼
              run_iadhore()
                    │
                    ▼
          read_iadhore_output()
```

---

## Installation

Full step-by-step installation instructions are in the
[README](https://github.com/lizhencmb/iadhoreR#installation). In brief:

1. Install [R](https://cran.r-project.org) (≥ 4.0)
2. Install [Miniconda](https://docs.conda.io/en/latest/miniconda.html) or
   [Mambaforge](https://github.com/conda-forge/miniforge#mambaforge)
3. Create the conda environment with all tools:

```bash
conda env create -f https://raw.githubusercontent.com/lizhencmb/iadhoreR/main/inst/conda/environment.yml
conda activate iadhoreR
```

   > **Windows users:** use WSL2 — see the
   > [README](https://github.com/lizhencmb/iadhoreR#windows-users-wsl2)
   > for setup instructions.

4. With the environment active, install iadhoreR from R:

```r
remotes::install_github("lizhencmb/iadhoreR", build_vignettes = TRUE)
```

### Verify the setup

```{r eval=FALSE}
library(iadhoreR)
check_tools()
#> External tool status:
#>   i-adhore     [OK]     /opt/conda/envs/iadhoreR/bin/i-adhore
#>   diamond      [OK]     /opt/conda/envs/iadhoreR/bin/diamond
#>   mcl          [OK]     /opt/conda/envs/iadhoreR/bin/mcl
#>   mcxload      [OK]     /opt/conda/envs/iadhoreR/bin/mcxload
#>   mcxdump      [OK]     /opt/conda/envs/iadhoreR/bin/mcxdump
#> All tools found. You are ready to use iadhoreR.
```

---

## Tutorial: *Arabidopsis* vs *Vitis* synteny

```{r eval=FALSE}
library(iadhoreR)

# All input files are bundled with the package
ath_gff   <- system.file("extdata", "annotation.selected_transcript.all_features.ath.gff3",
                          package = "iadhoreR")
vvi_gff   <- system.file("extdata", "annotation.selected_transcript.all_features.vvi.gff3",
                          package = "iadhoreR")
ath_fasta <- system.file("extdata", "proteome.selected_transcript.ath.fasta",
                          package = "iadhoreR")
vvi_fasta <- system.file("extdata", "proteome.selected_transcript.vvi.fasta",
                          package = "iadhoreR")

# Working directory for all outputs
work <- "iadhore_ath_vvi"
```

### Step 1 — Inspect the GFF files

Before parsing, use `inspect_gff()` to identify the correct feature type and
attribute key for your annotation files. This is especially useful when working
with a new GFF file you haven't seen before.

```{r eval=FALSE}
inspect_gff(ath_gff)
#> Feature types:
#>    exon     CDS    gene    mRNA  ...
#>
#> Format detected: GFF3
#>
#> Attribute keys per feature type:
#>   [mRNA]
#>   All keys: ID, Parent, Name, gene_id
#>   ID  e.g.: AT1G01010.1, AT1G01020.5, AT1G01030.1
```

The key information: use `feature_type = "mRNA"` and `id_attribute = "ID"`.

If you also have a FASTA file, `recommend_parse_gff()` goes further and
automatically finds the combination whose IDs best match the protein sequences:

```{r eval=FALSE}
rec <- recommend_parse_gff(ath_gff, ath_fasta)
#> FASTA: 48310 sequences
#>   e.g.: AT1G01010.1, AT1G01020.5, AT1G01030.1
#>
#> Best match: feature_type = "mRNA", id_attribute = "ID"
#>   GFF unique IDs : 27655
#>   FASTA          : 48310
#>   Matched        : 27655
#>   Chromosomes with matched genes (7): Chr1, Chr2, Chr3, Chr4, Chr5, ChrC, ChrM
```

> **Note:** The FASTA here contains all transcript isoforms (48,310), while the
> GFF uses one selected transcript per gene (27,655). `recommend_parse_gff()`
> automatically identifies the correct IDs to use.

### Step 2 — Parse GFF files into gene list files

i-ADHoRe requires one plain-text file per chromosome, listing gene IDs in
genomic order with their strand orientation. `parse_gff()` creates these files
from a GFF.

```{r eval=FALSE}
# Arabidopsis: bundled data contains Chr1 only
# For a full analysis use all nuclear chromosomes:
#   chromosomes = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5")
ath_lists <- parse_gff(
  ath_gff,
  output_dir   = file.path(work, "ath_lists"),
  genome_name  = "ath",
  feature_type = "mRNA",
  id_attribute = "ID",
  chromosomes  = "Chr1"
)
#> Wrote 1 gene list files to iadhore_ath_vvi/ath_lists

# Vitis: bundled data contains chr1 only
# For a full analysis use: chromosomes = paste0("chr", 1:19)
vvi_lists <- parse_gff(
  vvi_gff,
  output_dir   = file.path(work, "vvi_lists"),
  genome_name  = "vvi",
  feature_type = "mRNA",
  id_attribute = "ID",
  chromosomes  = "chr1"
)
#> Wrote 1 gene list files to iadhore_ath_vvi/vvi_lists
```

Each list file looks like:

```
AT1G01010.1+
AT1G01020.5-
AT1G01030.1+
...
```

The returned named vectors map chromosome IDs to file paths and are passed
directly to `write_iadhore_config()`.

### Step 3 — All-vs-all protein similarity search

i-ADHoRe needs to know which genes are homologous across and within genomes.
We use DIAMOND to run a fast all-vs-all blastp search. Passing both FASTAs
together produces cross-species hits alongside self-hits in a single run.

```{r eval=FALSE}
run_diamond(
  fasta_file  = c(ath_fasta, vvi_fasta),
  output_file = file.path(work, "all_vs_all.blast"),
  work_dir    = file.path(work, "diamond"),
  evalue      = 1e-5,
  threads     = 8,
  verbose     = FALSE
)
#> Combining 2 FASTA files...
#> Building DIAMOND database...
#> Running all-vs-all DIAMOND blastp...
#> DIAMOND search complete: iadhore_ath_vvi/all_vs_all.blast
```

### Step 4 — Cluster homologs into gene families with MCL

i-ADHoRe's `table_type=family` mode expects a two-column gene-to-family table.
`blast_to_families()` handles the full MCL pipeline: filtering BLAST hits,
building the graph with `mcxload`, clustering with `mcl`, and exporting with
`mcxdump`.

The `gene_list_files` argument is important: any gene present in the lists but
absent from BLAST hits (no detectable homolog) is assigned its own singleton
family, ensuring every gene has a family ID.

```{r eval=FALSE}
blast_to_families(
  blast_file      = file.path(work, "all_vs_all.blast"),
  output_file     = file.path(work, "families.txt"),
  gene_list_files = c(unname(ath_lists), unname(vvi_lists)),
  work_dir        = file.path(work, "mcl"),
  inflation       = 3.0,    # higher = smaller, tighter clusters
  evalue_cutoff   = 1e-5
)
#> Preparing ABC input from BLAST results...
#>   1842301 pairs written
#> Running mcxload...
#> Running MCL (inflation = 3)...
#> Running mcxdump...
#> Writing gene-to-family table...
#>   Added 8432 singleton families
#> Done: 54001 genes in 21876 families -> iadhore_ath_vvi/families.txt
```

The output file has two tab-separated columns — no header:

```
AT1G01010.1    FAM000001
AT1G01020.5    FAM000001
GSVIVT01000042001    FAM000001
AT1G01030.1    FAM000002
...
```

#### Choosing the inflation parameter

| Inflation | Effect |
|-----------|--------|
| 1.2 – 2.0 | Larger, looser families (more distant homologs grouped together) |
| 2.0 – 3.0 | Balanced — good default for most analyses |
| 3.0 – 5.0 | Smaller, tighter families (stricter ortholog grouping) |

For synteny detection, values between 2.0 and 3.0 work well for most plant
genomes.

### Step 5 — Write the i-ADHoRe configuration file

```{r eval=FALSE}
write_iadhore_config(
  genomes       = list(ath = ath_lists, vvi = vvi_lists),
  blast_table   = file.path(work, "families.txt"),
  table_type    = "family",
  output_path   = file.path(work, "iadhore_output"),
  file          = file.path(work, "config.ini"),

  # clustering parameters
  gap_size      = 30,   # max distance between genes in a colinear segment
  cluster_gap   = 35,   # max distance between segments to be grouped
  anchor_points = 3,    # min homologous gene pairs per segment
  q_value       = 0.75, # min linearity (R²) of a cluster
  prob_cutoff   = 0.001,

  number_of_threads = 8
)
```

#### Key parameters explained

| Parameter | What it controls | Typical range |
|-----------|-----------------|---------------|
| `gap_size` | Max pseudodistance between anchor points within a segment | 10 – 40 |
| `cluster_gap` | Max distance between base clusters to merge into a multiplicon | ≥ `gap_size` |
| `anchor_points` | Min anchor points for a valid segment | 3 – 6 |
| `q_value` | Minimum R² (linearity) of a cluster | 0.70 – 0.90 |
| `prob_cutoff` | Max probability of a cluster arising by chance | 0.001 – 0.01 |

Increase `anchor_points` and decrease `prob_cutoff` for stricter, higher-confidence
results. For whole-genome duplication studies, `anchor_points = 5` with
`gap_size = 30` is a common starting point.

### Step 6 — Run i-ADHoRe

```{r eval=FALSE}
run_iadhore(
  config_file = file.path(work, "config.ini"),
  verbose     = TRUE
)
#> i-ADHoRe 3.0 starting...
#> ...
#> i-ADHoRe finished.
```

> **Note:** i-ADHoRe sometimes exits with status 11 (segmentation fault) after
> writing all output. iadhoreR converts this to a warning rather than an error,
> since the output files are complete.

### Step 7 — Read and explore the results

```{r eval=FALSE}
results <- read_iadhore_output(file.path(work, "iadhore_output"))
names(results)
#> [1] "multiplicons"  "baseclusters"  "anchorpoints"  "segments"
#> [5] "genes"         "list_elements"
```

#### Multiplicons

The top-level syntenic regions. Each row is a pair of syntenic segments between
two genomic locations (`genome_x`/`list_x` vs `genome_y`/`list_y`).

```{r eval=FALSE}
head(results$multiplicons)
#>   id genome_x list_x parent genome_y list_y level number_of_anchorpoints
#> 1  1      ath   Chr1     NA      ath   Chr1     2                      5
#> 2  2      ath   Chr1     NA      ath   Chr2     2                      4
#> ...
#>   profile_length begin_x end_x begin_y end_y is_redundant
```

Key columns:
- `level` — multiplicon level; level 2 = pairwise, higher = profile-based
- `number_of_anchorpoints` — number of homologous gene pairs anchoring this region
- `is_redundant` — if TRUE, this region is a subset of a higher-level multiplicon

Count syntenic regions by species pair:

```{r eval=FALSE}
table(paste(results$multiplicons$genome_x, "vs", results$multiplicons$genome_y))
```

#### Anchor points

The individual homologous gene pairs that define each syntenic region.

```{r eval=FALSE}
head(results$anchorpoints)
#>   id multiplicon basecluster   gene_x            gene_y coord_x coord_y
#> 1  1           1           1 AT1G01010.1 GSVIVT01000001001       1       1
#> ...
#>   is_real_anchorpoint
```

To get all anchor gene pairs for a specific multiplicon:

```{r eval=FALSE}
mp_id <- 1
ap <- results$anchorpoints[results$anchorpoints$multiplicon == mp_id, ]
ap[, c("gene_x", "gene_y")]
```

#### Segments

The genomic boundaries of each syntenic segment.

```{r eval=FALSE}
head(results$segments)
#>   id multiplicon genome list first last order
#> 1  1           1    ath Chr1    10   45     1
#> 2  2           1    vvi chr1     3   31     2
```

#### Putting it together: genes in syntenic regions

To find all genes involved in any syntenic region:

```{r eval=FALSE}
# Gene IDs from anchor points
syntenic_genes <- unique(c(results$anchorpoints$gene_x,
                           results$anchorpoints$gene_y))
length(syntenic_genes)
```

---

### Step 8 — Summary statistics

`iadhore_summary()` prints a quick overview combining colinearity and
duplication information:

```{r eval=FALSE}
iadhore_summary(results)
#> === i-ADHoRe run summary ===
#>
#>   Total multiplicons     : 1482
#>   Non-redundant          : 1104
#>   Maximum level          : 5
#>
#> --- Colinear portions (genome level) ---
#>
#>   ath vs vvi : 31.42%
#>   vvi vs ath : 28.17%
#>
#> --- Duplicated portions (genome level) ---
#>
#>   ath : 47.31% duplicated
#>   vvi : 52.88% duplicated
```

The two underlying data frames are available separately if you need per-chromosome
detail:

```{r eval=FALSE}
# Colinear portions: one row per (genome, chromosome, other_genome)
cp <- colinear_portions(results)
head(cp)
#>   genome  list list_size other_genome colinear_count colinear_pct
#> 1    ath Chr1      7152          vvi           2301       32.17

# Genome-level roll-up
aggregate(cbind(colinear_count, list_size) ~ genome + other_genome,
          data = cp, FUN = sum)

# Duplicated portions: one row per (genome, chromosome, level)
mp <- multiplicated_portions(results)
head(mp)
#>   genome  list list_size level count   pct
#> 1    ath Chr1      7152     1  3891 54.41
#> 2    ath Chr1      7152     2  2651 37.07
#> 3    ath Chr1      7152     3   610  8.53

# Fraction of each genome covered at level >= 2
dup     <- mp[mp$level >= 2, ]
tot_cnt <- aggregate(count ~ genome, data = mp,  FUN = sum)
dup_cnt <- aggregate(count ~ genome, data = dup, FUN = sum)
merge(tot_cnt, dup_cnt, by = "genome", suffixes = c("_total", "_dup"))
```

---

### Step 9 — Homologous gene groups

`homolog_groups()` runs a union-find algorithm over all anchor pairs in all
multiplicons, grouping transitively homologous genes into discrete sets.  It
uses the full anchor point table (all levels), so if gene A pairs with B in a
level-2 multiplicon and B pairs with C in a level-3 multiplicon, all three end
up in the same group.

```{r eval=FALSE}
hg <- homolog_groups(results)
# One row per gene that appears in at least one anchor pair
head(hg)
#>   homolog_group group_size              gene genome  list coordinate
#> 1             1          3       AT1G01010.1    ath  Chr1          1
#> 2             1          3       AT3G01100.1    ath  Chr3        152
#> 3             1          3 GSVIVT01000042001    vvi  chr5         88
#>   remapped_coordinate orientation
#> 1                   1           +
#> 2                 150           +
#> 3                  87           +

# Size distribution of homolog groups
table(hg$group_size)
#>    2    3    4    5    6
#> 2140  612  184   56   12

# Largest groups
head(hg[order(-hg$group_size, hg$homolog_group), ])

# Export for downstream use (e.g. OrthoFinder comparison)
write.csv(hg, "homolog_groups.csv", row.names = FALSE)
```

> **Note:** Groups correspond to i-ADHoRe anchor pairs, not gene families.  A
> gene may be in a family (from MCL) but not in a homolog group if it has no
> syntenic partner.

---

### Step 10 — Visualisations

All three plot functions accept either a path string or the list returned by
`read_iadhore_output()`.

#### Whole-genome dot plot

`plot_dotplot()` places all chromosomes along each axis and draws one point per
anchor pair.  Points are coloured by multiplicon level (light → dark blue =
low → high level) by default.

```{r eval=FALSE}
# Inter-genomic: ath (x) vs vvi (y)
plot_dotplot(results,
             genome_x  = "ath",
             genome_y  = "vvi",
             chr_order = "natural")   # chr1 < chr2 < … < chr19

# Intra-genomic: vvi self-dot-plot (automatically mirrored across diagonal)
plot_dotplot(results,
             genome_x  = "vvi",
             genome_y  = "vvi",
             chr_order = "natural")
```

Key parameters:

| Parameter | Default | Effect |
|-----------|---------|--------|
| `chr_order` | `"size"` | `"natural"` sorts chr1 < chr2 < chr10; `"alpha"` alphabetically; `"size"` largest first |
| `real_only` | `TRUE` | Set to `FALSE` to also show non-real anchor points |
| `colour_by` | `"level"` | `"multiplicon"` or `"basecluster"` for discrete colours per block |
| `point_size` | `0.8` | Reduce (e.g. `0.3`) for dense whole-genome plots |

```{r eval=FALSE}
# Colour by multiplicon; smaller dots for a dense plot
plot_dotplot(results, genome_x = "vvi", genome_y = "vvi",
             colour_by = "multiplicon", point_size = 0.4)
```

#### Genome overview

`plot_genome_overview()` shows each chromosome as a horizontal baseline with
non-redundant multiplicon segments stacked above it.  Colour encodes level:
light grey = level 2, dark grey = levels 3–4, distinct colours ≥ level 5.
Segments are stacked with level 2 at the bottom so higher-level duplications
paint on top.

```{r eval=FALSE}
plot_genome_overview(results, genome = "ath", chr_order = "natural")
plot_genome_overview(results, genome = "vvi", chr_order = "natural")
```

| Parameter | Default | Effect |
|-----------|---------|--------|
| `chr_order` | `"natural"` | Same options as `plot_dotplot()` |
| `seg_height` | `0.3` | Height of each stacked rectangle |
| `band_gap` | `1.5` | Vertical spacing between chromosomes |

#### Multiplicon track diagram

`plot_multiplicon()` draws one horizontal gene track per segment of a single
multiplicon.  Genes are shown as directional arrows (→ forward, ← reverse).
Anchor pairs are connected by coloured bands; genes sharing the same homolog
group (transitively, across all ancestor multiplicons) share a colour.
Non-anchor genes are grey.

```{r eval=FALSE}
# Draw the first non-redundant multiplicon
first_mult <- results$multiplicons$id[results$multiplicons$is_redundant == 0][1]
plot_multiplicon(results, multiplicon_id = first_mult)

# Show gene IDs below each arrow (useful for small multiplicons)
plot_multiplicon(results, multiplicon_id = first_mult,
                 show_gene_ids = TRUE,
                 gene_id_cex   = 0.35)
```

For large multiplicons with many genes, widen the output device before plotting:

```{r eval=FALSE}
pdf("multiplicon_detail.pdf", width = 20, height = 5)
plot_multiplicon(results, multiplicon_id = first_mult, show_gene_ids = TRUE)
dev.off()
```

Key parameters:

| Parameter | Default | Effect |
|-----------|---------|--------|
| `real_only` | `TRUE` | Whether to restrict bands/colours to real anchor points |
| `band_alpha` | `0.35` | Transparency of connecting bands (0 = invisible, 1 = opaque) |
| `show_gene_ids` | `FALSE` | Print gene IDs rotated below each arrow |
| `gene_id_cex` | `0.4` | Text size for gene ID labels |

#### Saving plots to PDF

```{r eval=FALSE}
pdf("iadhore_plots.pdf", width = 10, height = 8)

plot_dotplot(results, genome_x = "ath", genome_y = "vvi", chr_order = "natural")
plot_dotplot(results, genome_x = "ath", genome_y = "ath", chr_order = "natural")
plot_genome_overview(results, genome = "ath")
plot_genome_overview(results, genome = "vvi")
plot_multiplicon(results, multiplicon_id = first_mult)

dev.off()
```

---

## Single-species analysis (intra-genomic duplications)

The workflow is identical — just provide one species and one FASTA:

```{r eval=FALSE}
ath_lists <- parse_gff(ath_gff,
  output_dir   = "ath_lists",
  genome_name  = "ath",
  feature_type = "mRNA",
  id_attribute = "ID",
  chromosomes  = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))

run_diamond(ath_fasta, output_file = "ath.blast", threads = 8)

blast_to_families("ath.blast",
  output_file     = "ath_families.txt",
  gene_list_files = unname(ath_lists))

write_iadhore_config(
  genomes     = list(ath = ath_lists),
  blast_table = "ath_families.txt",
  table_type  = "family",
  output_path = "ath_output",
  file        = "ath_config.ini")

run_iadhore("ath_config.ini")
results <- read_iadhore_output("ath_output")

# Statistics and plots work identically for single-species runs
iadhore_summary(results)
hg <- homolog_groups(results)

plot_dotplot(results, chr_order = "natural")          # intra-genomic dot plot
plot_genome_overview(results, genome = "ath")
```

---

## Working with your own data

### If you only have a BLAST/DIAMOND table (no MCL)

Use `parse_blast()` to filter the results and produce a gene-pair table for
i-ADHoRe's `table_type=pairs` mode:

```{r eval=FALSE}
parse_blast("mydata.blast",
  output_file   = "blast_pairs.txt",
  evalue_cutoff = 1e-5,
  min_pident    = 30)

write_iadhore_config(
  ...,
  blast_table = "blast_pairs.txt",
  table_type  = "pairs")
```

### Exploring a new GFF file

```{r eval=FALSE}
inspect_gff("my_annotation.gff3")
# Shows all feature types, attribute keys, and example IDs

recommend_parse_gff("my_annotation.gff3", "my_proteome.fasta")
# Automatically finds the best feature_type + id_attribute combination
```

### Writing gene lists manually

If your gene coordinates are already in R (e.g. from a database query), use
`write_gene_list()` directly:

```{r eval=FALSE}
write_gene_list(
  gene_ids = c("gene001", "gene002", "gene003"),
  strand   = c("+", "-", "+"),
  file     = "chr1.lst"
)
```

---

## Session information

```{r}
sessionInfo()
```
