---
title: "Getting Started with iadhoreR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with iadhoreR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**iadhoreR** is an R interface to
[i-ADHoRe 3.0](https://github.com/VIB-PSB/i-ADHoRe), a tool for detecting
collinear (syntenic) regions within and between genomes. It helps answer
questions like:

- Which chromosomal regions in species A are syntenic with regions in species B?
- Which genomic segments are remnants of ancient whole-genome duplications?
- Which segments descend from the same ancestral chromosome (putative ancestral
  regions, PARs)?

The package wraps the i-ADHoRe binary and the upstream tools it needs (DIAMOND
for homology search, MCL for gene-family clustering), and adds R helpers for
**preparing inputs**, **parsing every output table**, and **computing statistics
and plots** from the results.

This tutorial walks through a complete analysis using two bundled example
species (chromosome subsets included in the package; use your own full genomes
for real analyses):

- **ath** — *Arabidopsis thaliana*
- **vvi** — *Vitis vinifera*

The typical workflow has five stages:

```
GFF + FASTA                    i-ADHoRe config
    │                               │
    ▼                               ▼
parse_gff()   run_diamond()    write_iadhore_config()
    │               │               │
    │               ▼               │
    │         blast_to_families()   │
    │               │               │
    └───────────────┴───────────────┘
                    │
                    ▼
              run_iadhore()
                    │
                    ▼
          read_iadhore_output()  ──►  iadhore_summary(), find_pars(),
                                       plot_dotplot(), plot_genome_overview(),
                                       plot_multiplicon(), plot_par()
```

> Most code chunks below are shown for reference and are **not executed** when
> this vignette is built (they need the external binaries and full genomes).
> The input-preparation and output-parsing helpers, however, run on any machine
> with the package installed.

---

## Installation

Full step-by-step installation instructions are in the
[README](https://github.com/lizhencmb/iadhoreR#installation). In brief:

1. Install [R](https://cran.r-project.org) (≥ 4.0)
2. Install [Miniconda](https://docs.conda.io/en/latest/miniconda.html) or
   [Mambaforge](https://github.com/conda-forge/miniforge#mambaforge)
3. Create the conda environment with all tools:

```bash
conda env create -f https://raw.githubusercontent.com/lizhencmb/iadhoreR/main/inst/conda/environment.yml
conda activate iadhoreR
```

   > **Windows users:** use WSL2 — see the
   > [README](https://github.com/lizhencmb/iadhoreR#windows-users-wsl2)
   > for setup instructions.

4. With the environment active, install iadhoreR from R:

```r
remotes::install_github("lizhencmb/iadhoreR", build_vignettes = TRUE)
```

### Verify the setup

`check_tools()` reports whether each external binary can be found. If anything
is missing, `setup_instructions()` prints how to install it.

```{r eval=FALSE}
library(iadhoreR)

check_tools()
#> External tool status:
#>   i-adhore     [OK]     /opt/conda/envs/iadhoreR/bin/i-adhore
#>   diamond      [OK]     /opt/conda/envs/iadhoreR/bin/diamond
#>   mcl          [OK]     /opt/conda/envs/iadhoreR/bin/mcl
#>   mcxload      [OK]     /opt/conda/envs/iadhoreR/bin/mcxload
#>   mcxdump      [OK]     /opt/conda/envs/iadhoreR/bin/mcxdump
#> All tools found. You are ready to use iadhoreR.

setup_instructions()   # printed only when a tool is missing
```

The bundled example files used throughout this tutorial are located with
`system.file()`:

```{r eval=FALSE}
ath_gff   <- system.file("extdata", "annotation.selected_transcript.all_features.ath.gff3",
                         package = "iadhoreR")
vvi_gff   <- system.file("extdata", "annotation.selected_transcript.all_features.vvi.gff3",
                         package = "iadhoreR")
ath_fasta <- system.file("extdata", "proteome.selected_transcript.ath.fasta",
                         package = "iadhoreR")
vvi_fasta <- system.file("extdata", "proteome.selected_transcript.vvi.fasta",
                         package = "iadhoreR")

work <- "iadhore_ath_vvi"   # working directory for all outputs
```

---

## Stage 1 — Prepare the gene lists

i-ADHoRe requires, for each genome, one plain-text file per chromosome listing
gene IDs in genomic order with a strand tag (`AT1G01010+`). Those IDs must match
the IDs in your homology table, which in turn come from your protein FASTA. The
three helpers below take you from a raw GFF/GTF to those gene lists.

### 1a. Inspect a GFF — `inspect_gff()`

Before parsing, survey what feature types and attribute keys an annotation
contains. This runs live on the bundled file:

```{r}
library(iadhoreR)
ath_gff <- system.file("extdata",
  "annotation.selected_transcript.all_features.ath.gff3", package = "iadhoreR")

info <- inspect_gff(ath_gff)
```

The `mRNA` rows carry an `ID` like `AT1G01010.1` — exactly how the *Arabidopsis*
proteome headers are written — so `feature_type = "mRNA"`, `id_attribute = "ID"`.

### 1b. Match a GFF to a proteome — `recommend_parse_gff()`

You rarely have to choose those parameters by hand. `recommend_parse_gff()`
tries **every** feature-type / attribute-key combination present in the file and
reports the one whose IDs best overlap your protein FASTA:

```{r}
ath_fasta <- system.file("extdata", "proteome.selected_transcript.ath.fasta",
                         package = "iadhoreR")
rec <- recommend_parse_gff(ath_gff, ath_fasta)
```

```{r}
rec[c("feature_type", "id_attribute", "n_matched")]
```

Two things make this robust on real-world annotations:

- **All attribute keys are considered**, not a fixed shortlist. The matching ID
  often lives under a non-standard key — `protein_id` (Ensembl-style),
  `Source_ID`, `proteinId`/`transcriptId` (JGI-style) — and those are found
  automatically.
- **When the FASTA IDs genuinely don't match any attribute**, the function stops
  with an actionable message naming the attribute that corresponds to a FASTA
  header *sub-field*, instead of silently matching a handful of genes. That tells
  you exactly what to reconcile (rename FASTA headers, or rebuild the homology
  table with the GFF's IDs).

The right answer depends on how the proteome was keyed, not on a fixed
convention: the same call on a proteome that uses gene-level IDs would instead
recommend `feature_type = "gene"`.

### 1c. Write the gene lists — `parse_gff()`

Feed the recommended parameters in. `parse_gff()` writes one `.lst` file per
chromosome and returns a named vector mapping chromosome → file path, ready for
`write_iadhore_config()`.

```{r eval=FALSE}
ath_lists <- parse_gff(
  ath_gff,
  output_dir   = file.path(work, "ath_lists"),
  genome_name  = "ath",
  feature_type = rec$feature_type,   # "mRNA"
  id_attribute = rec$id_attribute,   # "ID"
  chromosomes  = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5")
)
#> Wrote 5 gene list files to iadhore_ath_vvi/ath_lists

vvi_lists <- parse_gff(
  vvi_gff,
  output_dir   = file.path(work, "vvi_lists"),
  genome_name  = "vvi",
  feature_type = "mRNA",
  id_attribute = "ID",
  chromosomes  = paste0("chr", 1:19)
)
```

Each list file looks like:

```
AT1G01010.1+
AT1G01020.5-
AT1G01030.1+
...
```

`parse_gff()` also handles two things that trip up naive parsing of real files:

- **One entry per gene.** When the chosen feature type repeats a gene across
  rows (typical for `CDS` or `exon`), only the first occurrence at the gene's
  lowest coordinate is kept — the list is not padded with duplicates.
- **Strand-safe.** Features whose strand is `.` (unknown) cannot be oriented in
  a gene list, so they are dropped with a warning rather than aborting the run.

### 1d. Writing gene lists manually — `write_gene_list()`

If your coordinates are already in R (e.g. from a database query), build the
lists directly:

```{r eval=FALSE}
write_gene_list(
  gene_ids = c("gene001", "gene002", "gene003"),
  strand   = c("+", "-", "+"),
  file     = "chr1.lst"
)
```

---

## Stage 2 — Build the homology table

i-ADHoRe needs to know which genes are homologous, supplied either as a
two-column **gene-pair** table (`table_type = "pairs"`) or a two-column
**gene-to-family** table (`table_type = "family"`).

### 2a. All-vs-all similarity search — `run_diamond()`

Pass both FASTAs together to get cross-species and within-species hits in one
run.

```{r eval=FALSE}
run_diamond(
  fasta_file     = c(ath_fasta, vvi_fasta),
  output_file    = file.path(work, "all_vs_all.blast"),
  work_dir       = file.path(work, "diamond"),
  evalue         = 1e-5,
  threads        = 8,
  more_sensitive = FALSE,
  verbose        = FALSE
)
#> Combining 2 FASTA files...
#> Building DIAMOND database...
#> Running all-vs-all DIAMOND blastp...
#> DIAMOND search complete: iadhore_ath_vvi/all_vs_all.blast
```

### 2b. Option A — gene pairs with `parse_blast()`

The simplest table: filter the DIAMOND hits and write the surviving query/subject
pairs. i-ADHoRe adds the reciprocal pair automatically.

```{r eval=FALSE}
parse_blast(
  blast_file    = file.path(work, "all_vs_all.blast"),
  output_file   = file.path(work, "blast_pairs.txt"),
  evalue_cutoff = 1e-5,
  min_pident    = 30      # optional percent-identity floor
)
#> Wrote 1843211 gene pairs to iadhore_ath_vvi/blast_pairs.txt
```

### 2c. Option B — gene families with `blast_to_families()`

For `table_type = "family"` mode, cluster homologs with MCL.
`blast_to_families()` runs the full pipeline: filter hits, build the graph with
`mcxload`, cluster with `mcl`, export with `mcxdump`.

The `gene_list_files` argument matters: any gene present in the lists but absent
from BLAST hits is given its own singleton family, so **every** gene has a family
ID.

```{r eval=FALSE}
blast_to_families(
  blast_file      = file.path(work, "all_vs_all.blast"),
  output_file     = file.path(work, "families.txt"),
  gene_list_files = c(unname(ath_lists), unname(vvi_lists)),
  work_dir        = file.path(work, "mcl"),
  inflation       = 3.0,    # higher = smaller, tighter clusters
  evalue_cutoff   = 1e-5
)
#> Preparing ABC input from BLAST results...
#> Running mcxload / MCL (inflation = 3) / mcxdump...
#>   Added 8432 singleton families
#> Done: 54001 genes in 21876 families -> iadhore_ath_vvi/families.txt
```

The output has two tab-separated columns, no header:

```
AT1G01010.1          FAM000001
AT1G01020.5          FAM000001
GSVIVT01000042001    FAM000001
AT1G01030.1          FAM000002
...
```

#### Choosing the inflation parameter

| Inflation | Effect |
|-----------|--------|
| 1.2 – 2.0 | Larger, looser families (more distant homologs grouped together) |
| 2.0 – 3.0 | Balanced — good default for most analyses |
| 3.0 – 5.0 | Smaller, tighter families (stricter ortholog grouping) |

For synteny detection, 2.0–3.0 works well for most plant genomes.

---

## Stage 3 — Configure and run i-ADHoRe

### 3a. Write the configuration — `write_iadhore_config()`

```{r eval=FALSE}
write_iadhore_config(
  genomes       = list(ath = ath_lists, vvi = vvi_lists),
  blast_table   = file.path(work, "families.txt"),
  table_type    = "family",                       # or "pairs"
  output_path   = file.path(work, "iadhore_output"),
  file          = file.path(work, "config.ini"),

  # clustering parameters
  cluster_type  = "colinear",
  gap_size      = 30,    # max distance between genes in a colinear segment
  cluster_gap   = 35,    # max distance between segments to be grouped
  anchor_points = 3,     # min homologous gene pairs per segment
  q_value       = 0.75,  # min linearity (R²) of a cluster
  prob_cutoff   = 0.001,

  number_of_threads = 8
)
```

#### Key parameters

| Parameter | What it controls | Typical range |
|-----------|-----------------|---------------|
| `gap_size` | Max pseudodistance between anchor points within a segment | 10 – 40 |
| `cluster_gap` | Max distance between base clusters to merge into a multiplicon | ≥ `gap_size` |
| `anchor_points` | Min anchor points for a valid segment | 3 – 6 |
| `q_value` | Minimum R² (linearity) of a cluster | 0.70 – 0.90 |
| `prob_cutoff` | Max probability of a cluster arising by chance | 0.001 – 0.01 |
| `multiple_hypothesis_correction` | `"FDR"`, `"bonferroni"`, or `"none"` | `"FDR"` |
| `level_2_only` | Restrict to pairwise (level-2) multiplicons | `FALSE` |

Increase `anchor_points` and decrease `prob_cutoff` for stricter, higher-confidence
results. For whole-genome duplication studies, `anchor_points = 5` with
`gap_size = 30` is a common starting point.

#### Cloud and hybrid modes

`cluster_type` also accepts `"cloud"` (degraded/diffuse homology) and `"hybrid"`
(colinear + cloud). These add their own parameters and produce two extra output
tables (see Stage 4):

```{r eval=FALSE}
write_iadhore_config(
  genomes      = list(ath = ath_lists, vvi = vvi_lists),
  blast_table  = file.path(work, "families.txt"),
  table_type   = "family",
  output_path  = file.path(work, "iadhore_cloud"),
  file         = file.path(work, "config_cloud.ini"),
  cluster_type = "hybrid",
  cloud_gap_size      = 20,
  cloud_cluster_gap   = 25,
  cloud_filter_method = "binomial"
)
```

### 3b. Run it — `run_iadhore()`

```{r eval=FALSE}
run_iadhore(config_file = file.path(work, "config.ini"), verbose = TRUE)
#> i-ADHoRe 3.0 starting...
#> ...
#> i-ADHoRe finished.
```

> **Note:** i-ADHoRe sometimes exits with status 11 (segmentation fault) after
> writing all output. iadhoreR converts this to a warning rather than an error,
> since the output files are complete.

---

## Stage 4 — Read the output

### `read_iadhore_output()` — everything at once

```{r eval=FALSE}
results <- read_iadhore_output(file.path(work, "iadhore_output"))
names(results)
#> [1] "multiplicons"  "baseclusters"  "anchorpoints"  "segments"
#> [5] "genes"         "list_elements"
#> # plus "clouds" and "cloud_anchorpoints" for cloud/hybrid runs
```

Each element is a data frame. You can also read any single table on its own with
the matching `read_*()` function:

| Function | Table | Key columns |
|----------|-------|-------------|
| `read_multiplicons()` | Syntenic region pairs | `id`, `level`, `number_of_anchorpoints`, `is_redundant` |
| `read_baseclusters()` | Diagonals within multiplicons | `id`, `multiplicon`, `orientation` |
| `read_anchorpoints()` | Homologous gene pairs | `gene_x`, `gene_y`, `is_real_anchorpoint` |
| `read_segments()` | Segment boundaries | `genome`, `list`, `first`, `last` |
| `read_genes()` | All genes with remapped coordinates | `genome`, `list`, `coordinate`, `orientation` |
| `read_list_elements()` | Gene ↔ segment membership | `segment`, `gene`, `position` |
| `read_clouds()` | Cloud regions (cloud/hybrid) | `cloud_density`, `dim_x`, `dim_y` |
| `read_cloud_anchorpoints()` | Cloud anchor pairs (cloud/hybrid) | `CloudID`, `gene_x`, `gene_y` |

```{r eval=FALSE}
# the multiplicons table: each row is a pair of syntenic segments
head(results$multiplicons)
#>   id genome_x list_x parent genome_y list_y level number_of_anchorpoints ...
#> 1  1      ath   Chr1     NA      ath   Chr1     2                      5 ...

# count syntenic regions per species pair
table(paste(results$multiplicons$genome_x, "vs", results$multiplicons$genome_y))

# all genes that anchor any syntenic region
syntenic_genes <- unique(c(results$anchorpoints$gene_x,
                           results$anchorpoints$gene_y))
```

---

## Stage 5 — Statistics and downstream analysis

### Quick overview — `iadhore_summary()`

```{r eval=FALSE}
iadhore_summary(results)
#> === i-ADHoRe run summary ===
#>   Total multiplicons     : 1482
#>   Non-redundant          : 1104
#>   Maximum level          : 5
#>
#> --- Colinear portions (genome level) ---
#>   ath vs vvi : 31.42%
#>   vvi vs ath : 28.17%
#>
#> --- Duplicated portions (genome level) ---
#>   ath : 47.31% duplicated
#>   vvi : 52.88% duplicated
```

### Per-chromosome detail

```{r eval=FALSE}
# Colinear portions: one row per (genome, chromosome, other_genome)
cp <- colinear_portions(results)
head(cp)
#>   genome  list list_size other_genome colinear_count colinear_pct
#> 1    ath  Chr1      7152          vvi           2301         32.17

# Duplicated portions: one row per (genome, chromosome, level)
mp <- multiplicated_portions(results)
head(mp)
#>   genome  list list_size level count   pct
#> 1    ath  Chr1      7152     2  2651  37.07
```

### Transitive homolog groups — `homolog_groups()`

Runs a union-find over all anchor pairs (all levels), grouping transitively
homologous genes into discrete sets:

```{r eval=FALSE}
hg <- homolog_groups(results)
head(hg)
#>   homolog_group group_size              gene genome  list coordinate ...
#> 1             1          3       AT1G01010.1    ath  Chr1          1 ...

table(hg$group_size)        # size distribution
write.csv(hg, "homolog_groups.csv", row.names = FALSE)
```

> Groups correspond to i-ADHoRe anchor pairs, not MCL families: a gene may be in
> a family but not in a homolog group if it has no syntenic partner.

### Putative ancestral regions — `find_pars()`

`find_pars()` clusters syntenic segments that share the same partner regions —
likely duplicated copies of one ancestral locus — following Jaillon *et al.*
(2009). It works directly on the i-ADHoRe output (no extra input needed) and
takes either a path or the `read_iadhore_output()` list.

```{r eval=FALSE}
pars <- find_pars(results, genome_x = "ath", genome_y = "vvi",
                  p_cutoff = 1e-10, r_cutoff = 0.3)
#> Found 31 PARs covering 88 significant segment pairs

pars$n_pars
head(pars$pairs)        # par_id, segment IDs and coordinate ranges, score
```

For an intra-genomic analysis, omit `genome_y` (it defaults to `genome_x`).

---

## Stage 6 — Visualisation

All plot functions accept either a path string or the `read_iadhore_output()`
list. Save several to one PDF by wrapping the calls in `pdf(...)` / `dev.off()`.

### Whole-genome dot plot — `plot_dotplot()`

One point per anchor pair, all chromosomes along each axis.

```{r eval=FALSE}
# inter-genomic: ath (x) vs vvi (y)
plot_dotplot(results, genome_x = "ath", genome_y = "vvi", chr_order = "natural")

# intra-genomic self dot plot (mirrored across the diagonal)
plot_dotplot(results, genome_x = "vvi", genome_y = "vvi",
             colour_by = "multiplicon", point_size = 0.4)

# overlay the PARs found above
plot_dotplot(results, genome_x = "ath", genome_y = "vvi",
             chr_order = "natural", highlight_pars = pars)
```

| Parameter | Default | Effect |
|-----------|---------|--------|
| `chr_order` | `"size"` | `"natural"` (chr1<chr2<chr10), `"alpha"`, or `"size"` (largest first) |
| `real_only` | `TRUE` | `FALSE` also shows non-real anchor points |
| `colour_by` | `"level"` | `"multiplicon"` or `"basecluster"` for discrete colours |
| `point_size` | `0.8` | reduce (e.g. `0.3`) for dense whole-genome plots |

### Genome overview — `plot_genome_overview()`

Each chromosome as a baseline with non-redundant multiplicon segments stacked
above it; colour encodes level.

```{r eval=FALSE}
plot_genome_overview(results, genome = "ath", chr_order = "natural")
plot_genome_overview(results, genome = "vvi", chr_order = "natural")
```

### Single multiplicon — `plot_multiplicon()`

One gene track per segment of a multiplicon; genes as directional arrows, anchor
pairs joined by coloured bands (shared colour = same homolog group).

```{r eval=FALSE}
first_mult <- results$multiplicons$id[results$multiplicons$is_redundant == 0][1]
plot_multiplicon(results, multiplicon_id = first_mult)

# label genes (best for small multiplicons / wide devices)
plot_multiplicon(results, multiplicon_id = first_mult,
                 show_gene_ids = TRUE, gene_id_cex = 0.35)
```

### PAR plots — `plot_par_overview()` and `plot_par()`

`plot_par_overview()` shows the segment-pair score matrix with the PAR clustering
dendrogram; `plot_par()` draws individual PARs on a dot-plot background.

```{r eval=FALSE}
plot_par_overview(pars, output_dir = file.path(work, "iadhore_output"))

plot_par(file.path(work, "iadhore_output"), pars)            # all PARs
plot_par(file.path(work, "iadhore_output"), pars, par_ids = 1:3)  # selected
```

---

## Exporting to MCScanX

If you want to view the synteny in [MCScanX](https://github.com/wyp1125/MCScanX)
or its downstream tools, convert the i-ADHoRe output to MCScanX's two input
formats:

```{r eval=FALSE}
# .collinearity file (synteny blocks), optionally with e-values from BLAST
write_mcscanx_collinearity(
  output_dir = file.path(work, "iadhore_output"),
  file       = file.path(work, "ath_vvi.collinearity"),
  blast_file = file.path(work, "all_vs_all.blast")  # optional
)

# .gff file (gene positions in MCScanX's chr/gene/start/end format)
write_mcscanx_gff(
  output_dir = file.path(work, "iadhore_output"),
  file       = file.path(work, "ath_vvi.gff")
)
```

---

## Single-species analysis (intra-genomic duplications)

The workflow is identical — provide one genome and its FASTA. This detects
ancient whole-genome-duplication blocks within the species.

```{r eval=FALSE}
ath_lists <- parse_gff(ath_gff, output_dir = "ath_lists", genome_name = "ath",
                       feature_type = "mRNA", id_attribute = "ID",
                       chromosomes = paste0("Chr", 1:5))

run_diamond(ath_fasta, output_file = "ath.blast", threads = 8)
blast_to_families("ath.blast", output_file = "ath_families.txt",
                  gene_list_files = unname(ath_lists))

write_iadhore_config(genomes = list(ath = ath_lists),
                     blast_table = "ath_families.txt", table_type = "family",
                     output_path = "ath_output", file = "ath_config.ini")
run_iadhore("ath_config.ini")

results <- read_iadhore_output("ath_output")
iadhore_summary(results)
pars <- find_pars(results, genome_x = "ath")    # intra-genomic PARs

plot_dotplot(results, genome_x = "ath", genome_y = "ath", chr_order = "natural")
plot_genome_overview(results, genome = "ath")
plot_par(results, pars)
```

---

## Function reference

| Stage | Functions |
|-------|-----------|
| Setup | `check_tools()`, `setup_instructions()` |
| Input prep | `inspect_gff()`, `recommend_parse_gff()`, `parse_gff()`, `write_gene_list()` |
| Homology | `run_diamond()`, `parse_blast()`, `blast_to_families()` |
| Configure / run | `write_iadhore_config()`, `run_iadhore()` |
| Read output | `read_iadhore_output()`, `read_multiplicons()`, `read_baseclusters()`, `read_anchorpoints()`, `read_segments()`, `read_genes()`, `read_list_elements()`, `read_clouds()`, `read_cloud_anchorpoints()` |
| Statistics | `iadhore_summary()`, `colinear_portions()`, `multiplicated_portions()`, `homolog_groups()`, `find_pars()` |
| Plots | `plot_dotplot()`, `plot_genome_overview()`, `plot_multiplicon()`, `plot_par_overview()`, `plot_par()` |
| Export | `write_mcscanx_collinearity()`, `write_mcscanx_gff()` |

Every function has a full help page — e.g. `?parse_gff` — listing all arguments
and return values.

---

## Session information

```{r}
sessionInfo()
```
